\chapter{Hilbert spaces}

\section{From vector to hilbert spaces}

	\subsection{Definition}

		\subsubsection{Classical mechanics}
		In classical mechanics the instantaneous state of a single particle is specified by the vector position $\vec{r}(t)$ and its momentum $\vec{p}(t)$ in the real vector spaces.
		In these vector spaces linearity holds: any linear combination of elements arbitrarily chosen inside the vector space $\mathcal{V}$ is still an element of the same vector space:
	
		$$\forall \vec{v}, \vec{w}\in\mathcal{V}\land\forall a, b\in\mathbb{R}: a\vec{v}+b\vec{w}\in\mathcal{V}$$

		Imposing that $a,b\in\mathbb{R}$ the discussion is restricted to real vector spaces.
		These have an operation called inner or scalar product that takes as input two vectors and gives as output a scalar.
		This is a real number for real vector spaces:

		$$(\vec{v}, \vec{w})\equiv \vec{v}\cdot\vec{w} = \vec{w}\cdot\vec{v}\in\mathbb{R}$$

		This operation is bi-linear:

		$$(a_1\vec{v}_1 + a_2\vec{v}_2)\cdot(b_1\vec{w}_1 + b_2\vec{w}_2) = \sum\limits_{i, k=1}^2 a_ib_k\vec{v}_i\vec{w}_k$$

		\subsubsection{Quantum mechanics}
		In quantum mechanics the state of a particle is instantaneously described by a quantum state $|\psi(t)\rangle$ that belongs to a hilbert space or $\mathcal{H}$, which provides a suitable generalization for the notion of vector speces.
		In a hilbert space any linear combination of quantum states is still inside the hilbert space:

		$$\forall |\psi\rangle,|\phi\rangle\in\mathcal{H}\land \alpha,\beta\in\mathbb{C}: \alpha|\psi\rangle + \beta|\phi\rangle\in\mathcal{H}$$

		A scalar product can be defined:

		$$\langle\psi|\phi\rangle = (\langle\phi|\psi\rangle)^*\in\mathbb{C}$$

		In hilbert spaces reversing the order of elements in the inner spaces leads to the complex conjugate result.
		The inner product is bi-linear:

		$$\begin{cases}|\xi\rangle \equiv \alpha_1|\psi_1\rangle + \alpha_2|\psi_2\rangle\\|\omega equiv \beta_1|\phi_1\rangle + \beta_2|\phi_2\end{cases}\Rightarrow \langle\xi|\omega\rangle = \sum\limits_{i=1}^2\alpha_i^*\beta_i\langle \psi_i|\phi_i\rangle$$

	\subsection{Complete orthonormal bases}
	
		\subsubsection{Standard vector spaces}
		In standard vector spaces orthonormal basis vectors are a set form of elements which are mutually orthogonal and have unitary norm:

		$$\vec{e}_i\cdot\vec{e}_j = \delta_{ij}$$

		Where $\delta_{ij}$ is the kronecker-delta and is defined as:

		$$\delta_{ij} = \begin{cases}1\qquad i = j\\ 0\qquad i\neq j\end{cases}$$

		It also has the property that:

		$$\sum\limits_j\delta_{ij}A_j = A_i$$

		A set of $N$ versors is said to form a complete orthonormal basis of the vector space $\mathcal{V}$ if for any vector $\vec{v}\in\mathcal{V}$ there exists a unique set of $N$ real coefficient $\lambda_1, \dots, \lambda_N$ which enable to express $\vec{v}$ as a linear combination of the basis:

		$$\vec{v} = \sum\limits_{k=1}^N\lambda_k\vec{e}_k$$

		$N$ is de dimension of the vector spece.
		The set of coefficients is called the coordinates of the vector $\vec{v}$ in the given complete orthonormal basis $\{e_k\}_{k = 1, \dots, N}$.

		\subsubsection{Hilbert space}
		A set of quantum states is defined to be a complete orthonormal basis of the Hilbert space $\mathcal{H}$ if: 
		\begin{multicols}{2}
			\begin{itemize}
				\item They are mutually orthogonal.
				\item They have unit norm.
				\item Any state can be written as linear combination of them.
			\end{itemize}
		\end{multicols}

		The elements of a complete orthonormal basis of a Hilbert space may form an infinite and dens set.
		Let's consider the set of position quantum states $|x\rangle$.
		Clearly two position can differ by an infinitesimal amount, therefore a continuous index $x$ is needed to label them.
		Two position states are said to obey the orthonormally condition if if holds:

		$$\langle \vec{x}|\vec{y}\rangle = \delta(\vec{x}-\vec{y}_)$$

		$\delta(\vec{x}-\vec{y})$ denotes the dirac-delta:

		$$\int d^3\vec{y} A(y)\delta(\vec{x}-\vec{y}) = A(\vec{x})$$

		The fact that position states form a basis of $\mathcal{H}$ expresses the fact that any quantum state in the Hilbert space can be obtained from a linear combination of the position states:

		$$|\psi\rangle = \int d^3 \vec{x}\phi(\vec(x))|vec{x}\rangle\qquad \phi(\vec{x})\in\mathbb{C}$$

		The complex function $\phi(x)$ is called the wave function and can be regarded as a dense and infinite set of complex coefficients.
		Therefore Hilbert spaces are infinite dimension vector spaces.

	\subsection{Operators}
	Operators are defined by their action on the elements of the vector space:

	$$\vec{w} = \hat{O}\vec{v}$$

	In particular $\hat{O}$ is linear if:

	$$\hat{O}(\alpha_1\vec{v}_1 + \alpha_2\vec{v}_2) = \alpha_1\hat{O}\vec{v}_1 + \alpha_2\hat{O}\vec{v}_2$$

	Once a basis of a $N$-dimensional real vector space is defined, then each linear operator can be assigned a $N\times N$ matrix, through the representation of the operator in the specific basis:

	$$\vec{w} = \hat{O}\vec{v}\rightarrow \vec{e}_j\cdot\vec{w} = \sum\limits_{i=1}^N(\vec{e}_i\cdot\vec{v})\vec{e}_j\cdot\hat{O}\vec{e}_i = \sum\limits_{i=1}^NO_{ji}v_j$$



	
